---
title: "README"
output:
 github_document:
 pandoc_args: ["--wrap=none"]
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#Load the dataset
baseball_data <- read.csv('./Data/baseball_1985_2016.csv')

#Load necessary packages 
library(knitr)
library(ggplot2)
library(dplyr)
library(corrplot)
library(tidyr)
library(ggpubr)
library(stats)
library(car)
library(kableExtra)
#install.packages("ggcorrplot")  # if not already installed
library(ggcorrplot)
```

## Introduction 

As college students, we are constantly busy with classes, homeworks, and exams. We are all very involved in extracurricular activities outside of classes, which takes up a lot of time. However, in the free time that we do have, we love to get together with our friends and watch sports games. Whether it is the fall semester or the spring semester, there is always a sport being played (football in the fall, basketball in the winter, baseball in the spring). Since this project is being completed in the spring season, we decided to focus our attention on baseball. 

Specifically, we are interested in learning more about what influences a baseball player’s salary. Is it their batting average? Their experience? Their age? The team they play for? We will analyze these variables, plus many others, and how they influence a player’s salary. In this case, we are interested in assessing the following question: What best influences a hitters’ baseball player’s salary? 

To start, we will understand the data by reading literature about impacts on a baseball player’s salary. This will allow us to absorb the material we are trying to learn more about before actually starting to import and analyze the data into an open source data software. Then, we will provide some descriptive and numeric statistics of the dataset we collected. This will help provide an overview of the important features of the variables. Next, we will do some exploratory data analysis, analyzing any correlations between the independent variables of interest and the dependent variable. We will utilize plots and graphs to help us visualize the data. Finally, we will use machine learning models to see what the best way to identify patterns, make predictions, and improve its performance over time. 

We will calculate the batting average by dividing each player’s total hits per season by their total at-bat. This batting average is different because a hit is a batter reaching first base. It is a good starting point for evaluating a player’s offensive skills. A higher batting average suggests a batter is more consistent in making solid contact with the ball and getting on base via hits. 

## Data Organization

Data was collected using a website called Baseball Reference, which has baseball history and statistics for Major League Baseball. It is the complete source for current and historical baseball players, teams, scores, and leaders. Specifically, our dataset focuses on statistics from 1985, showing individuals’ performance, age, and other relevant information for a singular season. 

According to the Bureau Labor of Statistics, salary trends change over time. With free agencies and salary arbitration, salaries escalate. In addition, TV revenues fuel salary expansion, as the League has more money to work with. In order to understand how much a team can spend on its players, it is important to understand the distribution of their money. For the owners, player compensation looks like this: Major League Player Compensation + Benefit Plan Costs + Postseason Share Payments + Minor League Signing Bonuses (not including associated tax) + Minor League Salaries And Benefits. 
	
So, in order to comprehend the salary of an individual player, we must take this distribution into account. Salaries to the players themselves are comparable to salaries in the entertainment industry: star performers disproportionately earn higher amounts. These higher amounts are influenced by performance, seniority, and market size, with performance being the dominant factor. 

```{r, include=FALSE}
#Keeping only selected variables to evaluate
hitters_data <- baseball_data %>%
  select(playerid, yearid, hits_p, game_p, stint, atbat_b, hr_b, rbi_b, salary, birthyear)  

#Log salary to scale the data
hitters_data$log_Salary <- log(hitters_data$salary)

#Replace original salary data with logged salary
hitters_data <- hitters_data %>%
  filter(salary > 0) %>%
  mutate(salary = log(salary))

#Delete duplicate row of logged salary
hitters_data <- hitters_data[, -9]
```


After filtering the data by removing the irrelevant attributes to our analysis, these are the attributes included in our data, as well as their data type:

```{r, include=FALSE}
colnames(hitters_data) <- c("Player_ID", "Year_ID", "Hits", "Games Played", "Stint", "At Bat", "Homeruns", "RBI", "Birth Year", "Logged_Salary")

#Feature info table 
feature_info <- data.frame(
  Feature = colnames(hitters_data),
  Description = c("Player ID", "Year ID", "Number of hits per season", "Games played per season","How many years player has been on team", "Number of times at bat per season", "Home runs per season","Runs Batted In per season", "Birth Year", "Yearly salary in dollars, logged" 
  ),
  Scale_Type = c("Character", "Ordinal", "Interval", "Interval", "Interval", "Interval", "Interval", "Interval", "Interval", "Ratio"
  )
)
```


### Feature Descriptions for Baseball Data

```{r, echo=FALSE}
# Display the table nicely
kable(feature_info)
```

Some of the players in the dataset were pitchers and not hitters, so we had to remove rows with NA values for hits in order to only look at the hitters. 

```{r, echo=FALSE}
# Remove any NA or NaN value
hitters_data <- hitters_data[!is.na(hitters_data$Hits), ]

# Function to summarize data with key statistics, excluding Player_ID from summary only
get_summary_stats <- function(data) {
  data %>%
    select(where(is.numeric), -Year_ID) %>%  # keep only numeric vars, EXCEPT Year_ID
    summarise(across(
      everything(),
      list(
        Count = ~sum(!is.na(.)),
        Mean = ~mean(., na.rm = TRUE),
        Median = ~median(., na.rm = TRUE),
        SD = ~sd(., na.rm = TRUE),
        Min = ~min(., na.rm = TRUE),
        Max = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )) %>%
    pivot_longer(
      cols = everything(),
      names_to = c("Variable", "Statistic"),
      names_pattern = "^(.*)_(Count|Mean|Median|SD|Min|Max)$",
      values_to = "Value"
    ) %>%
    pivot_wider(names_from = Statistic, values_from = Value)
}

# Summary Table
summary_baseball <- get_summary_stats(hitters_data)

kable(summary_baseball, caption = "Detailed Summary Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

## Exploratory Data Analysis

### Distribution of Variables 

```{r, echo=FALSE}
# Histogram for Hits
ggplot(hitters_data, aes(x = Hits)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Hits", x = "Hits", y = "Frequency") +
  theme_minimal()

# Density plot for Hits
ggplot(hitters_data, aes(x = Hits)) +
  geom_density(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Density of Hits", x = "Hits", y = "Density") +
  theme_minimal()
```

The distribution of Hits appears right-skewed, with a larger number of players accumulating fewer hits. The peak of the histogram is centered around a moderate value, suggesting that most players have a relatively average number of hits. There is a tail extending towards higher values, which indicates that there are a small number of players with many hits. The spread of the data, particularly with the long tail on the higher end, suggests that there are outliers or a small group of players who have significantly more hits than the majority.

```{r, echo=FALSE}
# Histogram for Birth Year 
ggplot(hitters_data, aes(x = `Birth Year`)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Birth Year", x = "Birth Year", y = "Frequency") +
  theme_minimal()

# Density plot for Birth Year 
ggplot(hitters_data, aes(x = `Birth Year`)) +
  geom_density(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Density of Birth Year", x = "Birth Year", y = "Density") +
  theme_minimal()
```

The distribution of Birth Year shows the distribution of players' birth years, with clear peaks around the 1940s, 1960s, and 1980s. The highest frequencies occur for players born in the 1960s and 1980s, with a smaller peak for those born in the 1940s. Fewer players were born in the earlier (1930s) or later years (late 1990s and beyond), indicating that most players fall into these key birth-year ranges, likely reflecting the active playing years of those players in the dataset.

```{r, echo=FALSE}
# Histogram for At Bat
ggplot(hitters_data, aes(x = `At Bat`)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of At Bat", x = "At Bat", y = "Frequency") +
  theme_minimal()

# Density plot for At Bat
ggplot(hitters_data, aes(x = `At Bat`)) +
  geom_density(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Density of At Bat", x = "At Bat", y = "Density") +
  theme_minimal()
```

The At Bat variable shows that most players have very few at-bats, with the distribution being heavily skewed towards lower values. Only a small subset of players have a significant number of at-bats, indicating a few high-usage players in contrast to many with minimal playing time.

```{r, echo=FALSE}
# Histogram for log_Salary
ggplot(hitters_data, aes(x = Logged_Salary)) +
  geom_histogram(binwidth = 0.1, fill = "lightgreen", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Logged Salary", x = "Log Salary", y = "Frequency") +
  theme_minimal()

# Density plot for log_Salary
ggplot(hitters_data, aes(x = Logged_Salary)) +
  geom_density(fill = "lightgreen", color = "black", alpha = 0.7) +
  labs(title = "Density of Logged Salary", x = "Log Salary", y = "Density") +
  theme_minimal()

```

The distribution of logged salary shows a right-skewed distribution, with most players having salaries concentrated between log values of 12 and 13. This suggests that the majority of players earn salaries within a moderate range, while a small number of players have significantly higher salaries, creating a long right tail. This indicates the presence of outliers, typical of professional sports where a few players earn much more than the majority.

The variables are not normally distributed, so we cannot run tests that assume normality, such as the Shapiro-Wilk test or Q-Q plots.
Instead, we would use the Pearson Correlation test.

### Pearson Correlation heatmap

Before running any exploratory analysis, we decided to see if there are any strong correlations that stand out for our independent variables of interest and our dependent variable, logged salary. According to the heatmap scale, the more correlated the variables are, the darker the color will be. We also have the actual correlations between the variables presented in each box. 


```{r, echo=FALSE}
#Compute Pearson correlation
numeric_filtered_data <- hitters_data[sapply(hitters_data, is.numeric)]
cor_matrix <- cor(numeric_filtered_data, method = "pearson")

ggcorrplot(cor_matrix, lab = TRUE, type = "lower", title = "Pearson Correlation Heatmap")

```

Surprisingly, we found that there are not many strong correlations between the variables of interest and our dependent variable. Even though the strongest correlation between an independent variable and salary is Year ID at .46, this is not helpful to our interpretation of how strongly the data is correlated since we are including this variable to represent that this is not time series data. Therefore, this correlation is not important. The next strongest correlation is for the independent variable, hits, with a correlation coefficient	of .26. This means that out of all the independent variables we looked at, hits are most likely to predict the outcome of a players’ salary. This is as expected, as the more hits a player has, the more likely they are going to get paid higher salaries, since their overall performance is high. The correlation of 0.21 between Birth Year and logged salary is moderate and positive, meaning that players who are older (i.e., having a higher birth year number) tend to have higher logged salaries. This might reflect the effect of experience and seniority, as older players may earn higher salaries due to their experience in the league.The correlation between At Bat and logged salary is 0.11, which indicates a weak positive relationship. This implies that while there is a slight tendency for players with more at-bats to have a higher logged salary, the relationship is not substantial.

### Check statistical significance of variables

```{r, echo=FALSE}
model <- lm(Logged_Salary ~ Hits + `Games Played` + Stint + `At Bat` + Homeruns + RBI + `Birth Year`, data = hitters_data)

model_summary <- summary(model)

# Extract coefficients
coef_table <- as.data.frame(model_summary$coefficients)

# Optionally rename the columns for clarity
colnames(coef_table) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)")

# Load knitr for pretty printing
library(knitr)

# Print nicely formatted table
kable(coef_table, caption = "Linear Regression Results for Logged_Salary")

```

In order to test if the correlations between these variables and salary were significant, we ran hypothesis tests on each of the variables depicted in the heatmap. We ran a t-test to determine if the p-values were small enough for each variable to be statistically significant. After running these tests, Hits, Games Played, and Birth Year all have highly significant p-values (well below 0.001), meaning these variables strongly influence logged salary. Stint is also significant (p < 0.01), but its effect is negative. At Bat and Homeruns have non-significant p-values (above 0.05), meaning they do not meaningfully contribute to the model. RBI is marginally significant with a p-value close to 0.1, suggesting a weak but possibly important relationship with salary. 

Our analysis in the next section will include only the variables that had a statistically significant effect on salary. 

### Check Assumptions 

Because we ran statistical tests, we need to run tests to check for linearity, homoscedasicity, and multicollinearity. We found above that the variables are not normally distributed, so we are running non-parametric tests. 

```{r, echo=FALSE}
# Reduced model with only statistically significant variables
model_sig <- lm(Logged_Salary ~ Hits + `Games Played` + Stint + `Birth Year`, data = hitters_data)

# Check linearity 
plot(model_sig, which = 1)  # Residuals vs Fitted

# Check homoscedasticity
library(lmtest)
bptest(model_sig)  # Breusch-Pagan test

# Check multicollinearity
vif(model_sig)

```

Linearity:
The Residuals vs. Fitted plot shows the residuals (the differences between the observed and predicted values) on the y-axis and the fitted values (predicted values) on the x-axis. The plot displays a random scatter of points around the horizontal line at zero, indicating that the model is generally well-fitted. The slightly decreasing trend in the residuals, however, suggests a mild non-linear pattern that the model might not be capturing perfectly. This could be a signal to consider potential improvements, such as adding non-linear terms or interactions. Overall, the residuals seem evenly distributed around zero, implying no severe heteroskedasticity or major model misspecifications.

Breusch-Pagan Test:
Interpretation: The very small p-value (< 0.05) strongly suggests that the residuals do not have constant variance — i.e., heteroscedasticity is present.
Why it matters: This can lead to inefficient estimates and invalid p-values, so it's a critical issue.
Fix: Use robust standard errors when reporting inference

Variance Inflation Factor:
Interpretation: All VIFs are very close to 1, indicating no multicollinearity.
No action needed — this assumption is met.

### Scatterplots of statistically significant variables

```{r, echo=FALSE}
library(ggplot2)

# Scatterplot: Hits vs Logged_Salary
ggplot(hitters_data, aes(x = Hits, y = Logged_Salary)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Hits vs Logged Salary", x = "Hits", y = "Log(Salary)") +
  theme_minimal()

# Scatterplot: Hits vs Games Played
ggplot(hitters_data, aes(x = `Games Played`, y = Logged_Salary)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Games Played vs Logged Salary", x = "Games Played", y = "Log(Salary)") +
  theme_minimal()

# Scatterplot: Hits vs Stint
ggplot(hitters_data, aes(x = Stint, y = Logged_Salary)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Stint vs Logged Salary", x = "Stint", y = "Log(Salary)") +
  theme_minimal()

# Scatterplot: Hits vs Birth Year
ggplot(hitters_data, aes(x = `Birth Year`, y = Logged_Salary)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Birth Year vs Logged Salary", x = "Birth Year", y = "Log(Salary)") +
  theme_minimal()

```



## Conclusions 

The goal of our analysis was to analyze if there was a correlation between the number of hits an individual MLB player has during a season and the logged salary. It found that 


## Main Observations 



## Future Directions

This information that we have extracted from the dataset can help determine how much players need to be contributing to their team in order to increase their salaries. It would now be interesting to look at how the other half of baseball, the outfield, affects individual players salaries. This would differ since it would have to be specific to position in the outfield, for example a pitcher and a first baseman could not be judged on the same scale. There would need to be multiple forms for each position across teams and use variables like pitches, catches, forced outs, and others in order to look at the impact on their salary. 

## More Information 

Youtube video links here 
