---
title: "README"
output: html_document
date: "2025-04-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#Load the dataset
baseball_data <- read.csv('baseball_1985_2016.csv')

#Load necessary packages 
library(knitr)
library(ggplot2)
library(dplyr)
library(corrplot)
library(tidyr)
library(ggpubr)
library(stats)
library(car)
library(kableExtra)
```

## Data Organization
```{r, include=FALSE}
#Keeping only selected variables to evaluate
filtered_data <- baseball_data %>%
  select(playerid, yearid, hits_p, salary)  

#Log salary to scale the data
filtered_data$log_Salary <- log(filtered_data$salary)

#Replace original salary data with logged salary
filtered_data <- filtered_data %>%
  filter(salary > 0) %>%
  mutate(salary = log(salary))

#Delete duplicate row of logged salary
filtered_data <- filtered_data[, -4]

```


After filtering the data by removing the irrelevant attributes to our analysis, these are the attributes included in our data, as well as their data type:

```{r, include=FALSE}
colnames(filtered_data) <- c("Player_ID", "Year_ID", "Hits", "Logged_Salary")

#Feature info table
feature_info <- data.frame(
  Feature = colnames(filtered_data),
  Description = c("Player ID", "Year ID", "Number of hits per season", "Yearly salary in dollars, logged" 
  ),
  Scale_Type = c("Character", "Ordinal", "Interval", "Interval"
  )
)
```


### Feature Descriptions for Baseball Data

```{r, echo=FALSE}
# Display the table nicely
kable(feature_info)
```

Some of the players in the dataset were pitchers and not hitters, so we had to remove rows with NA values for hits in order to only look at the hitters. 

```{r, echo=FALSE}
# Remove any NA or NaN value
filtered_data <- filtered_data[!is.na(filtered_data$Hits), ]

# Function to summarize data with key statistics, excluding Player_ID from summary only
get_summary_stats <- function(data) {
  data %>%
    select(where(is.numeric), -Year_ID) %>%  # keep only numeric vars, EXCEPT Year_ID
    summarise(across(
      everything(),
      list(
        Count = ~sum(!is.na(.)),
        Mean = ~mean(., na.rm = TRUE),
        Median = ~median(., na.rm = TRUE),
        SD = ~sd(., na.rm = TRUE),
        Min = ~min(., na.rm = TRUE),
        Max = ~max(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )) %>%
    pivot_longer(everything(),
                 names_to = c("Variable", "Statistic"),
                 names_sep = "_",
                 values_to = "Value") %>%
    pivot_wider(names_from = Statistic, values_from = Value)
}

# Summary Table
summary_baseball <- get_summary_stats(filtered_data)

kable(summary_baseball, caption = "Detailed Summary Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```
## Exploratory Data Analysis
```{r, echo=FALSE}
# Histogram for Hits
ggplot(filtered_data, aes(x = Hits)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Hits", x = "Hits", y = "Frequency") +
  theme_minimal()

# Histogram for log_Salary
ggplot(filtered_data, aes(x = Logged_Salary)) +
  geom_histogram(binwidth = 0.1, fill = "lightgreen", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Logged Salary", x = "Log Salary", y = "Frequency") +
  theme_minimal()


# Density plot for Hits
ggplot(filtered_data, aes(x = Hits)) +
  geom_density(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Density of Hits", x = "Hits", y = "Density") +
  theme_minimal()

# Density plot for log_Salary
ggplot(filtered_data, aes(x = Logged_Salary)) +
  geom_density(fill = "lightgreen", color = "black", alpha = 0.7) +
  labs(title = "Density of Logged Salary", x = "Log Salary", y = "Density") +
  theme_minimal()
```
###


